Designing Types for R, Empirically

### 2020-09-14 paper/main.tex
updated plots, finalizing edits

% Kaggle
 \newcommand{\PERCFAILEDASSERTIONSKAGGLE}{2.14\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSKAGGLENOUNDEF}{0.42\%\xspace}
 \newcommand{\PROPFUNSWITHFAILEDASSERTKAGGLE}{15.98\%\xspace}
 \newcommand{\NUMTOTALASSERTIONSKAGGLE}{1,9038,496\xspace}
 \newcommand{\NUMASSERTEDFUNCTIONSKAGGLE}{970\xspace}

@@

\begin{abstract}
The R programming language is widely used in a variety of domains.  It was
designed to favor an interactive style of programming with minimal syntactic
and conceptual overhead. This design is well suited to interactive data
analysis, but a bad fit for tools such as compilers or program analyzers
which must generate native code or catch programming errors.  In particular,
R has no type annotations, all operations are dynamically checked at
run-time. The starting point for our work are the twin questions, \emph{what
  expressive power is needed to accurately type R code?} and \emph{which
  type system is the R community willing to adopt?} Both questions are
difficult to answer without actually experimenting with a type system. The
goal of this paper is to provide data that can feed into that design
process. To this end, we perform a large corpus analysis to gain insights in
the degree of polymorphism exhibited by idiomatic R code and explore
potential benefits that the R community could accrue from a simple type
system.  As a starting point, we infer type signatures for \NUMFUNCTIONS
functions from \NUMPACKAGES packages among the most widely used open source
R libraries.
\end{abstract} \maketitle

@@

Our community builds, improves, and reasons about programming languages. To
make design decisions that benefit users, we need to understand our target
language as well as the real-world needs it answers. Often, we can appeal to
our intuition, as many languages are intended for general purpose
programming tasks.  Unfortunately, intuition may fail when looking at
domain-specific languages as these languages were designed for a specific
group of users to solve very specific needs. This is the case of the data
science language R.

@@

Many of the design decisions that gave us R were intended to foster an
interactive and exploratory programming style. These include, to name a few,
the lack of type annotations, the ability to use syntactic shortcut, and the
automatic conversion between data types. While these choices have led to a
language with a low barrier to entry---many data science educational
programs do not teach R itself but simply introduce some of its key
libraries---they have also created a language where errors can go
undetected.

Retrofitting a type system to the R language would increase our assurance in
the result of data analysis. But, we are faced with two challenges. First,
it is unclear what would be the \emph{right} type system for a language as
baroque as R. For example, one of the most popular data type, the
\code{data.frame}, is manipulated through reflective operations---a data
frame is a table whose columns can be added or removed on the fly.  Second,
but just as crucially, designing a type system that will be adopted would
require overcoming some prejudices and educating large numbers of users.

@@

This paper focuses on a simpler problem than an entire type system, instead,
we limit our scope to giving types to function signatures. For this, we
design a simple type language, one that matches the R data types but omits
features such as parametric polymorphism and subtyping between user-defined
data types. We then extract type signature from execution traces of a corpus
of widely used libraries.  This allows us to see how far one can get with
the simple type language and identify limitations of the design. We validate
the robustness of the extracted type signature by implementing a contact
checker that weaves the types around their respective functions, and use a
large number of clients of the target packages for validation. The contract
system can be used by the R community to experiment with type signatures and
replace ad hoc error checking code.

@@

Our tools are open source and publicly available on GitHub (link anonymize),
all results in this paper are reproducible and will be submitted for
artifact evaluation should this paper be accepted.
%
\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Dynamic programming languages such as Racket, JavaScript and PHP have been
extended post-facto with various static type systems.  In each case, the
type system was carefully engineered to match the salient characteristics of
the host languages and to foster a particular programming style. For
example, Typed Racket emphasizes functional programming and support the
migration from untyped to fully typed code~\cite{tf-popl08},
Hack~\cite{hack13} and TypeScript~\cite{BAT14} focus on typing
object-oriented features of PHP and JavaScript, respectively. They allow to
intersperse typed and untyped code in a fine-grained manner.  Other work
looks at retrofitting types onto Lua~\cite{typedlua2014}.

@@

In this paper, we will attempt infer type signature for functions. 
We begin by introducing some relevant concepts.

The basic building blocks of R are vectors of primitives.
Characters (strings), integers, logicals (booleans), doubles, complex numbers, and raw bytes make up R's primitive values.
When we say that R is {\it vectorized}, we mean that primitive values are {\it always} vectors.
For example, in R one might see \code{2.5} and think it to be a scalar double, but it is in actuality a unit-length vector containing a single double.
Vectors can also be explicitly constructed using the vector constructor \code{c(...)}, 
For example, \code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer literals are denoted by an \code{L}).

The main object of our work is R functions, which have a number of quirks.
Arguments to functions in R can be assigned default values (which can be arbitrary expressions!), and R functions can support variable argument lists, and functions can be called by specifying argument names explicitly in the call.
To illustrate all of this in a single example, consider:

@@

\noindent
This function four formal parameters, \k x, \k{...}, \k y and \k z. Argument
\k x can be bound positionally or passed by name. The vararg argument,
\k{...}, is always passed positionally. The remaining two arguments must be
passed by name.  Arguments \k y and \k z have default values, in the case of
\k z this is a constant, but \k{y}'s default value is an expression that
depends on the value of \k z. The body of the function will add parameters
\k x and \k y with either the scalar 0 or the result of concatenating the
varargs into a primitive vector. The function \k{missing} tests if an
argument was explicitly passed. The following are some valid invocations
of \k f:

\begin{lstlisting}
 > f(1)    
 [1] 2             % a double vector, y is 0, ... is missing
 > f(1,2) 
 [1] 4             % a double vector, y is 0, ... is 2
 > f(1,2,3)
 [1] 4 5           % a double vector, y is 0, ... is 2, 3
 > f(2,3,x=1)
 [1] 4 5           % a double vector, y is 0, ... is 2, 3
 > f(x=1, y= 1)
 [1] 2             % a double vector, y is 1, ... is missing
 > f(x=1, z= 1)
 numeric(0)        % a double vector of length 0, y is NULL
 > f(1L,2L,y=1L)
 [1] 4             % an integer vector, y is integer 1, ... is integer 2
 > f(1, y=c(1,2))
 [1] 1 2           % a double vetor, y is 1, 2, ... is missing
\end{lstlisting}

@@

\begin{lstlisting}
 > c1 <- complex(re=1, im=2)
 > c2 <- complex(re=2, im=1)
 > f(c1)
 [1] 1             % a double vector, x is complex, y is double
 > f(c1,y=c2)
 [1] 2+1i          % a complex vector, x and y are complex
 > l1 <- list(1,2)
 > l2 <- list(c1,c2)
 > f(l1) 
 [1] 1             % a double vector, x is a list of doubles
 > f(l1,y=l2)
 [[1]]             % a list of complex, x is a list of doubles
 [1] 1+2i          %                      y is a list of complex

@@

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language, which are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R, but in many
context a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}.

All vectorized data types have a distinguished missing value denoted by \code{NA}. 
The default type of \code{NA} is \k{logical}. 
We can see that \code{typeof(NA)=="logical"}, but NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.
In addition to \code{NA}, R has a \code{NULL} value, and to understand why R has both a \code{NULL} value and \code{NA} values it is useful to remember that R is primarily designed for statistics and data science.
In experimental data science, it is useful to have a notion of a ``missing observation'', which is often NaN (not a number, or similar) in, e.g., MATLAB and Python.
Since vectors are unityped in R, it {\it must} have a ``NaN'' value for each primitive type, hence the \code{NA}s, while there is only a single \code{NULL} null-typed value in the language to encompass all other cases.

@@

With one exception all vectorized data types are monomorphic, the exception
is the \k{list} type which can hold values of any other type including
\k{list}. For all monomorphic data types, attempting to store a value of a
different type will cause a conversion. Either the value is converted to the
type of vector, or the vector is converted to the type of the value.

Over the years, programmers have found the need for a richer type structure
and have added {\it attributes}. The best way to think of attributes is as
an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.  The addition of
attributes lets programmers extend the set of types by tagging data with
user-defined attributes. For example, one could define a vector of four
values, \code{x<-c(1,2,3,4)} and then attach the attribute \k{dim} with a
pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
arithmetic functions will treat \k{x} as a 2x2 matrix. Another attribute
that can be set is the \k{class}.  This attribute can be bound to a list of
class names. For instance, \code{class(x)<-"human"}, set the class of \k{x}
to be \k{human}.  Attributes are thus used for object-oriented
programming. The S3 object system support single dispatch on the class of
the first argument of a function, whereas the S4 object system allows
multiple dispatch (on all arguments). Some of the most widely used data
type, such as data frames, leverage attributes. A data frame, for instance,
is a list of vectors with a class and a column name attribute.

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types listed in Figure~\ref{rtypes} for reference.

@@

This section sets out to propose a candidate design for a type language to
describe the arguments and return values of functions in R.  The goal is not
to get a final design but rather a starting point for an iterative process.
Fig.~\ref{types} presents our type language. An early design choice was to
stay close to the R language and only depart in small, hopefully,
non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
variable length argument list. The rest of this section details and
motivates our design.

@@

\paragraph{NA values} In R, a value can be \emph{not available}, written
\NA. Each of basic types has its specific \NA. Thus there is an \int \NA as
well as a double \NA, and they are different.  The type system allows to
distinguish between vectors that may contain {\NA}s(written \NAVEC\int) and
those who are guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be
treated as a vector that may {\NA}s.  We do not allow scalar values to be
\NA and the type \raw does not allow {\NA}s.

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{~|~} \T_n$.

\paragraph{Nullables} 
The \nul type is inhabited by a singleton \code{NULL} value often used as a
sentinel or a stand in for arguments with no value.  To capture this
behavior, we introduce a nullable type \code{?}\T.  The difference between
the \NA and \code{NULL} values is that the latter cannot be stored in
vectors.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related: a vector converts to a list with
\code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). The list type is parameterize, \LIST\T, when the list is
heterogeneous \T defaults to \any.

\paragraph{Classes} 
R has more than one notion of type. Values can be attributed, one important
attribute is the class of a value.  A class is a list of names that are used
to mimic object-oriented programming. The type system includes class types
written \CLASS{\ID_1,\dots,\ID_n}.

@@

While this task is seemingly simple, the details and their proverbial devil
are surprisingly tricky to get right and to scale to large, long running,
programs.  To avoid starting from scratch, our implementation reuses \rdt,
an open source dynamic analysis framework for R~\cite{oopsla19} which
consists of an instrumented R Virtual Machine based on GNU-R version
3.5.0. The framework exposes hooks in the interpreter to which user defined
callbacks can be attached. These hooks include function entry and exit,
method dispatch for the S3 and S4 object systems, the longjumps used by the
interpreter to implement non-local exit, creation and forcing of promises,
variable definition, value creation, mutation and garbage collection.

@@

\noindent
To obtain these raw types, we make use of R's C FFI and use low-level
machinery to collect information from the R language run-time.  Types are
conmpleted during post-processing, and rely on the detailed information made
available by these low-level reflection mechanisms.

@@

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are termed
missing. This occurs when a function was called with too few arguments and
no default value was specified for missing arguments.  We record a
\k{missing} type for such argument.  There are two obvious ways to deal with
missing arguments: type them as \any or type them as some unit type.  As we
are performing a dynamic analysis, we conservatively type them as
\LIST{\any}.

\paragraph{Non-local returns}
When a function exits with a longjump, there is no return value to speak
of. To ensure call traces are valid when a longjump occurs, we intercept the
unwinding process and record a special \k{jumped} return type for function
returns that are skipped.

@@

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
\k{special\_exit}, \k{promise\_force\_entry}, and
\k{prom\-ise\_force\_exit}.  The function-related callbacks are used mainly
for bookkeeping: the analysis is notified that a construct has been entered
by pushing the call onto a stack.  The calls themselves store a trace
object, it is that object that holds the type information. As R can perform
single or multiple dispatch on function arguments depending on their class,
the relevant information is kept by the {\tt \_entry} variants.

@@

Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, immutable lists, and remove \nul type to replace them with
nullables. For performance reasons, we bound lists to no more than five
types in their union, above that they become \LIST{any}.  Functions are alos
limited to twenty arguments. Higher-order functions are approximated by \any
to \any.

@@

One way to validate the inferred type annotations is to check that different
programs using one of the functions for which a type signature has been inferred
respect the signature. For this purpose, we developed \contractr, an R package
that allows one to decorate function with assertions. We use it to insert type
checks of the arguments and return values.

@@

In terms of usability, our checker is enabled automatically whenever a package
is loaded in R. So, a simple invocation of \code{library(contractr)} causes
contracts to be injected in all packages. On loading, \contractr scans all
packages in the user's workspace and inserts contracts in functions for which
type signatures are available. Then \contractr sets up package load hooks which
are executed when new packages are loaded. Furthermore, \contractr automatically
removes contracts from all functions and restores them to their original state
when it is unloaded. The type signatures can be provided externally, in a file,
thus avoiding the need to change the source code of checked packages. Type
declarations can also be written directly in the adjunct function comments using
\roxygen annotation tags. \contractr introduces a new use \code{@type} tag. When
a package these type signatures are extracted and stored in the appropriate
location. The following is an example of a type annotation using \roxygen:

@@

The \contractr provides an extensive API that allows users to explicitly insert
contracts during interactive development by supplying the type signature as a
string, selectively disable contract checking and inspecting type assertions.

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely whether a value is a: null, environment, vector
or scalar. Other properties require an inspection of the contents of a
value: the absence of \NA or the type of the elements of list. Union types
require checking for each of the alternatives. We do not check higher-order functions in this version of the tool. Checking
them could be done by dynamically wrapping functions with checks. The usage
of higher-order functions remains limited in R. When argument values are wrapped in promises (this is not always the case
due to compiler optimizations), in order to retain the non-strict semantics
of R, the expression held in the promise is wrapped in a call to the type
checker, and type checking is delayed until the promise is forced. This
leads to corner cases such that the type checking of a function may happen
after that function has returned.

Return values require care as well.  Functions return the last expression
they evaluate, thus to avoid having to analyze the code of the called
function the checker will register a callback on the exit hook. The hook is
executed in the function call's environment. Another wrinkle is due to
longjumps which causes active function calls on the stack to be
discarded. When they are discarded, their exit hooks are called.  But, these
function do not have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as the
return value for calls that are discarded. The exit hook does not call the
type-checker if it see the sentinel.

@@

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
code (C/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of
the dots reflects the project's size in lines of code including both R and
native code\footnote{Lines of source code reported excludes comments and
  blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
expression code coverage in percents and the y-axis gives the number of
reverse dependencies in log scale. Dotted lines indicate means. Packages
with over \PackageSizeOutierRnd lines of code are annotated.

@@

These packages come from the Comprehensive R Archive Network
(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
\AllNativeCodeRnd R and native code respectively. Unlike other open code
repositories such as GitHub, CRAN is a curated repository. Each submitted
package must abide to a number of well-formedness rules that are automatically
checked asserting certain quality. Most relevant for this work is that all of
the runnable code (including code snippets from examples and vignettes) is
tested and only a successfully running package is admitted in the archive.

>>>>

% Kaggle
 \newcommand{\PERCFAILEDASSERTIONSKAGGLE}{2.14\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSKAGGLENOUNDEF}{0.42\%\xspace}
 \newcommand{\PROPFUNSWITHFAILEDASSERTKAGGLE}{15.98\%\xspace}
 \newcommand{\NUMTOTALASSERTIONSKAGGLE}{19,038,496\xspace}
 \newcommand{\NUMASSERTEDFUNCTIONSKAGGLE}{970\xspace}

@@

\begin{abstract}
The R programming language is widely used in a variety of domains.  It was
designed to favor an interactive style of programming with minimal syntactic
and conceptual overhead. This design is well suited to interactive data
analysis, but a bad fit for tools such as compilers or program analyzers
which must generate native code or catch programming errors.  In particular,
R has no type annotations, and all operations are dynamically checked at
run-time. The starting point for our work are the twin questions: \emph{what
  expressive power is needed to accurately type R code?} and \emph{which
  type system is the R community willing to adopt?} Both questions are
difficult to answer without actually experimenting with a type system. The
goal of this paper is to provide data that can feed into that design
process. To this end, we perform a large corpus analysis to gain insights in
the degree of polymorphism exhibited by idiomatic R code and explore
potential benefits that the R community could accrue from a simple type
system.  As a starting point, we infer type signatures for \NUMFUNCTIONS
functions from \NUMPACKAGES packages among the most widely used open source
R libraries. We then conduct an evaluation on \NUMPKGSEVAL clients of these packages, as well as on end-user code found on the Kaggle competition website.
\end{abstract} \maketitle

@@

Our community builds, improves, and reasons about programming languages. To
make design decisions that benefit users, we need to understand our target
language as well as the real-world needs it answers. Often, we can appeal to
our intuition, as many languages are intended for general purpose
programming tasks.  Unfortunately, intuition may fail when looking at
domain-specific languages designed for a specific
group of users to solve very specific needs. This is the case of the data
science language R.

@@

Many of the design decisions that gave us R were intended to foster an
interactive and exploratory programming style. These include, to name a few,
the lack of type annotations, the ability to use syntactic shortcuts, and the
automatic conversion between data types. While these choices have led to a
language with a low barrier to entry---many data science educational
programs do not teach R itself but simply introduce some of its key
libraries---they have also created a language where errors can go
undetected.

Retrofitting a type system to the R language would increase our assurance in
the result of data analysis, but we are faced with two challenges. First,
it is unclear what would be the \emph{right} type system for a language as
baroque as R. For example, one of the most popular data type, the
\code{data.frame}, is manipulated through reflective operations---a data
frame is a table whose columns can be added or removed on the fly.  Second,
but just as crucially, designing a type system that will be adopted would
require overcoming some prejudices and educating large numbers of users.

@@

This paper focuses on a simpler problem than an entire type system, instead,
we limit our scope to giving types to function signatures. For this, we
design a simple type language, one that matches the R data types but omits
features such as parametric polymorphism and subtyping between user-defined
data types. We then extract type signature from execution traces of a corpus
of widely used libraries.  This allows us to see how far one can get with
the simple type language and identify limitations of the design. We validate
the robustness of the extracted type signature by implementing a contact
checker that weaves the types around their respective functions, and use a
both a large number of clients of the target packages as well as end-user code for validation. The contract
system can be used by the R community to experiment with type signatures and
replace ad hoc error checking code.

@@

Our tools are open source and publicly available on GitHub (link anonymize),
all results in this paper are reproducible and were submitted for
artifact evaluation.
%
\section{Background} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Dynamic programming languages such as Racket, JavaScript and PHP have been
extended post-facto with various static type systems.  In each case, the
type system was carefully engineered to match the salient characteristics of
the host languages and to foster a particular programming style. For
example, Typed Racket emphasizes functional programming and support the
migration from untyped to fully typed code~\cite{tf-popl08},
Hack~\cite{hack13} and TypeScript~\cite{BAT14} focus on typing
object-oriented features of PHP and JavaScript, respectively. They allow to
intersperse typed and untyped code in a fine-grained manner.  Other work
looks at retrofitting types onto Lua~\cite{typedlua2014} which account for the myriad ways Lua programmers use tables.

@@

The basic building blocks of R are vectors of primitives (primitives in R are characters (strings), integers, logicals (booleans), doubles, complex numbers, and raw bytes).
When we say that R is {\it vectorized}, we mean that primitive values are {\it always} vectors.
For example, in R one might see the literal \code{2.5} and think it to be a scalar double, but it is in actuality a unit-length vector containing a single double.
Vectors can also be explicitly constructed using the vector constructor \code{c(...)}, 
For example, \code{c(1L, 2L, 3L)} creates a vector of three integers (in R, integer literals are denoted by an \code{L}).
To illustrate that scalars are vectors, in R \code{(c(2.5) == 2.5) == TRUE}.

The main object of our work is R functions, and functions in R have a number of quirks.
Arguments can be assigned default values (which can be arbitrary expressions!), and functions can support variable argument lists, and functions can be called by specifying argument names explicitly in the call.
To illustrate all of this in a single example, consider:

@@

\noindent
This function has four formal parameters, \k x, \k{...}, \k y and \k z. Argument
\k x can be bound positionally or passed by name. The vararg argument,
\k{...}, is always passed positionally. The remaining two arguments must be
passed by name as they are preceded by the \k{...} parameter.  Arguments \k y and \k z have default values, in the case of
\k z this is a constant, but \k{y}'s default value is an expression that
depends on the value of \k z. The body of the function will add parameters
\k x and \k y with either the scalar 0 or the result of concatenating the
varargs into a primitive vector. The function \k{missing} tests if an
argument was explicitly passed. The following are some valid invocations
of \k f:

\begin{lstlisting}
 > f(1)    
   [1] 2             # a double vector, y is 0, ... is missing
 > f(1,2) 
   [1] 4             # a double vector, y is 0, ... is 2
 > f(1,2,3)
   [1] 4 5           # a double vector, y is 0, ... is 2, 3
 > f(2,3,x=1)
   [1] 4 5           # a double vector, y is 0, ... is 2, 3
 > f(x=1, y= 1)
   [1] 2             # a double vector, y is 1, ... is missing
 > f(x=1, z= 1)
   numeric(0)        # a double vector of length 0, y is NULL
 > f(1L,2L,y=1L)
   [1] 4             # an integer vector, y is integer 1, ... is integer 2
 > f(1, y=c(1,2))
   [1] 1 2           # a double vetor, y is 1, 2, ... is missing
\end{lstlisting}

@@

\begin{lstlisting}
 > c1 <- complex(re=1, im=2)
 > c2 <- complex(re=2, im=1)
 > f(c1)
   [1] 1             # a double vector, x is complex, y is double
 > f(c1,y=c2)
   [1] 2+1i          # a complex vector, x and y are complex
 > l1 <- list(1,2)
 > l2 <- list(c1,c2)
 > f(l1) 
   [1] 1             # a double vector, x is a list of doubles
 > f(l1,y=l2)
   [[1]]             # a list of complex, x is a list of doubles
   [1] 1+2i          #                      y is a list of complex

@@

R has a builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{rtypes} lists all of the builtin types that are
provided by the language; these are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R, but in many
contexts a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}.

All vectorized data types have a distinguished missing value denoted by \code{NA} (for ``not available''). 
The default type of \code{NA} is \k{logical}. 
We can see that \code{typeof(NA)=="logical"}, but \k{NA} inhabits every type: \code{typeof(c(1,NA)[2])} \k{==} \k{"double"}.
In addition to \code{NA}, R has a \code{NULL} value, and to understand why R has both a \code{NULL} value and \code{NA} values it is useful to remember that R is primarily designed for statistics and data science.
In experimental data science, it is useful to have a notion of a ``missing observation'', which is often NaN (not a number, or similar) in, e.g., MATLAB and Python.
Since vectors are unityped in R, it {\it must} have a ``NaN'' value for {\it each} primitive type.
There single \code{NULL} null-typed value in the language to encompass all other cases.
In a sense, you can think of \code{NA} as representing a missing data point, and \code{NULL} as a missing data {\it set}.

@@

With one exception, all vectorized data types are monomorphic.
The exception is the \k{list} type which can hold values of any other type including
\k{list}. For all monomorphic data types, attempting to store a value of a
different type will cause a conversion. Either the value is converted to the
type of the vector, or the vector is converted to the type of the value.

Over the years, programmers have found the need for a richer type structure
and have added {\it attributes}. The best way to think of attributes is as
an optional map from names to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.  The addition of
attributes lets programmers extend the set of types by tagging data with
user-defined attributes. For example, one could define a vector of four
values, \code{x<-c(1,2,3,4)} and then attach the attribute \k{dim} with a
pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From that point,
arithmetic functions will treat \k{x} as a 2x2 matrix.

Another attribute
that can be set is the \k{class}.  This attribute can be bound to a list of
class names. For instance, \code{class(x)<-"human"} will set the class of \k{x}
to be \k{human}.  The main usage of attributes is for object-oriented
programming. There are three object-orientation frameworks in R: S3, S4, and R5. 
The S3 object system support single dispatch on the class of
the first argument of a function, whereas the S4 object system allows
multiple dispatch (on all arguments). R5 allows for users to define objects
in a more imperative style. Some of the most widely used data
types leverage attributes, e.g., data frames and matrices. A data frame, for instance,
is a list of vectors with a class and a column name attribute, and matrices are vectors with a \k{dims} attribute.

Scalar data types include the distinguished \k{NULL} value, instances of classes written using the S4 object system, closures, and environments.  The implementation of R has a number of other
types listed in Figure~\ref{rtypes} for reference.

@@

In this section, we set out to propose a candidate design for a type language to
describe the arguments and return values of R functions.  The goal is not
to propose a final design, but rather to propose a starting point for an iterative design process.
Fig.~\ref{types} presents our type language. An early design choice was to
stay close to the R language and only depart in small and hopefully
non-controversial ways.  Functions types have the form \FUN{A_1, \dots,
  A_n}\T where each $\A_i$ argument is either a type \T or \VARGS, a
variable length argument list. The rest of this section details and
motivates our design.

@@

\paragraph{NA values} In R, a value can be \emph{not available}, written
\NA. Each of basic types has its specific \NA. Thus there is an \int \NA as
well as a \dbl \NA, and they are different.  The type system allows one to
distinguish between vectors that may contain {\NA}s (written \NAVEC\int) and
those who are guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be
treated as a vector that may {\NA}s.  We do not allow scalar values to be
\NA and the type \raw does not allow {\NA}s.

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{~|~} ... \k{~|~} \T_n$.

\paragraph{Nullables} 
The \nul type is inhabited by a singleton \code{NULL} value often used as a
sentinel or a stand in for arguments with no value.  To capture this
behavior, we introduce a nullable type \code{?}\T.  The functional difference between
the \NA and \code{NULL} values is that the latter cannot be stored in
vectors.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related: a vector converts to a list with
\code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). The list type is parameterize: \LIST\T.

\paragraph{Classes} 
R has more than one notion of type. As we mentioned, values can be attributed, and a very important
attribute is the class of a value.  A class is a list of names that are used
to mimic object-oriented programming. The type system includes class types
written \CLASS{\ID_1,\dots,\ID_n}.

@@

While this task is seemingly simple, the details and their proverbial devil
are surprisingly tricky to get right and to scale to large, long running
programs.  To avoid starting from scratch, our implementation reuses \rdt,
an open source dynamic analysis framework for R~\cite{oopsla19} which
consists of an instrumented R Virtual Machine based on GNU-R version
3.5.0. The framework exposes hooks in the interpreter to which user defined
callbacks can be attached. These hooks include function entry and exit,
method dispatch for the S3 and S4 object systems, the longjumps used by the
interpreter to implement non-local exit, creation and forcing of promises,
variable definition, value creation, mutation and garbage collection.

@@

\noindent
To obtain these raw types, we make use of R's C FFI and use low-level
machinery to collect information from the R language run-time.  Types are
completed during post-processing, and rely on the detailed information made
available by these low-level reflection mechanisms.

@@

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are termed
missing. This occurs when a function was called with too few arguments and
no default value was specified for missing arguments.  We record a
\k{missing} type for such argument.  There are two obvious ways to deal with
missing arguments: type them as \any or type them as some unit type.  As we
are performing a dynamic analysis, we conservatively type them as
\any.

\paragraph{Non-local returns}
When a function exits with a longjump, there is no return value to speak
of. To ensure call traces are valid when a longjump occurs, we intercept the
unwinding process and record a special \k{jumped} return type for function
returns that are skipped. As we cannot be sure of the intended return value, 
these \k{jumped} become \any types.

@@

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry}, \k{closure\_exit},
\k{builtin\_entry}, \k{builtin\_exit}, \k{special\_entry},
\k{special\_exit}, \k{promise\_force\_entry}, and
\k{prom\-ise\_force\_exit}.  The function-related callbacks are used mainly
for bookkeeping: the analysis is notified that a construct has been entered
by pushing the call onto a stack.  The calls themselves store a trace
object that holds the type information. As R can perform
single or multiple dispatch on function arguments depending on their class,
the relevant information is kept by the {\tt \_entry} variants.

@@

Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, immutable lists, and remove \nul type to replace them with
nullables.  Functions are
limited to twenty arguments as a simplifying assumption. Higher-order functions are approximated by \any
to \any by \typetracer.

@@

One way to validate an inferred function type is to check that it is respected in a different program which uses the function.
For this purpose, we developed \contractr, an R package
that allows one to decorate function with assertions. We use it to insert type
checking of arguments and return values.

@@

In terms of usability, our checker is enabled automatically whenever a package
is loaded in R. So, a simple invocation of \code{library(contractr)} causes
contracts to be injected in all packages. On loading, \contractr scans all
packages in the user's workspace and inserts contracts in functions for which
type signatures are available. Then \contractr sets up package load hooks which
are executed when new packages are loaded. Furthermore, \contractr automatically
removes contracts from all functions and restores them to their original state
when it is unloaded. The type signatures can be provided in an external file,
thus avoiding the need to change the source code of checked packages. Type
declarations can also be written directly in the adjunct function comments using
\roxygen annotation tags. \contractr introduces a new use \code{@type} tag. When
a package these type signatures are extracted and stored in the appropriate
location. The following is an example of a type annotation using \roxygen:

@@

\contractr provides an extensive API that allows users to explicitly insert
contracts during interactive development by supplying the type signature as a
string, selectively disable contract checking and inspecting type assertions.

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely whether a value is a null, environment, vector
or scalar. Other properties require an inspection of the contents of a
value, such as checking for the absence of \NA or the type of the elements of list. Union types
require checking for each of the alternatives. As the usage
of higher-order functions remains limited in R, we do not check higher-order functions in this version of the tool. Checking
them could be done by dynamically wrapping functions with checks.  When argument values are wrapped in promises (this is not always the case
due to compiler optimizations), in order to retain the non-strict semantics
of R, the expression held in the promise is wrapped in a call to the type
checker, and type checking is delayed until the promise is forced. This
leads to corner cases such that the type checking of a function may happen
after that function has returned.

Return values require care as well.  Functions return the last expression
they evaluate, thus to avoid having to analyze the code of the called
function the checker will register a callback on the exit hook. The hook is
executed in the function call's environment. Another wrinkle is due to
longjumps which causes active function calls on the stack to be
discarded. When they are discarded, their exit hooks are called but they do not have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as the
return value for calls that are discarded. The exit hook does not call the
type-checker if it see the sentinel.

@@

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
code (C/Fortran). Figure~\ref{fig:corpus} shows these packages: the size of
the dots reflects the project's size in lines of code including both R and
native code\footnote{Lines of source code reported excludes comments and
  blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
expression code coverage as a percentage and the y-axis gives the number of
reverse dependencies in on log scale. Dotted lines indicate means. Packages
with over \PackageSizeOutierRnd lines of code are annotated.

@@

These packages come from the Comprehensive R Archive Network
(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
code with over \AllCranRnd packages\footnote{CRAN receives about 6 new package
  submissions a day~\cite{Ligges2017}} containing over \AllRCodeRnd and
\AllNativeCodeRnd lines of R and native code respectively. Unlike other open code
repositories such as GitHub, CRAN is a curated repository where each submitted
package must abide by a number of well-formedness rules that are automatically
checked to assess package quality. 
Notably, CRAN packages must have a set of {\it runnable} example, test, and vignette code which showcase package functionality.
The code is run by CRAN, and only a successfully running package is admitted to the archive.
